---
title: "Applying Probability-Ratio to Published Data"
author: "Matthew Jackson"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include = F, message=F, warning=F, echo=F, scipen=999)
library(here)
library(IgAScores)
library(tidyverse)
library(ggpubr)
library(ggsci)
library(pheatmap)
library(cowplot)
library(knitr)
library(broom)
library(phyloseq)
library(data.table)
library(phenoDist)
library(ggrepel)
library(ggbeeswarm)

theme_set(theme_pubr())

```

# Overview

The aim of this script is to apply the Probability Ratio to a published dataset. Huus et al. applied the Kau ratio to study binding of bacteria by IgA in under nourished mice relative to controls, finding they do not develop IgA binding of Lactobacillus in the under nourished state but do conventionally. Here, we carry out a reanalysis of a subset of these data kindly provided by the authors.

# Load in the data

```{r load_abunds_paper_code}
#following code taken from Huus paper
parse_taxonomy_simple = function(char.vec) {
  ranks = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")
  rv = strsplit(char.vec, ";")[[1]]
  rv = c(rv, rep(NA, length(ranks) - length(rv)))
  names(rv) = ranks
  rv
}
#import data
Table = import_qiime(otufilename = here("Huus_Data","feature-table-for-R-2.txt"), mapfilename = here("Huus_Data","sample-metadata.txt"), parseFunction = parse_taxonomy_simple)

#Remove halomonas and shewanella #and others
taxotu <- data.frame(OTU = row.names(otu_table(Table)), tax_table(Table))
halotus <- subset(taxotu, Domain == "Archaea"  | Genus == "Halomonas" | Genus == "Shewanella" | Family == "mitochondria" | Class == "chloroplast")
halotus <- as.character(halotus[["OTU"]])
Table <- prune_taxa(setdiff(row.names(otu_table(Table)), halotus), Table)

#remove singletons
Table <- prune_taxa(taxa_sums(Table) > 0, Table)

#prune taxa with a very low fractional abundance, ie 0.005% out of total taxa
number <- taxa_sums(Table)/sum(taxa_sums(Table))
Igf <- prune_taxa(number > 0.00005, Table)


```

```{r data_prep}
#load in the metadata, remove EE to keep balanced design
meta <- read_delim(here("Huus_Data","sample-metadata.txt"),"\t") %>% rename("samplename"=1) %>% filter(!Diet=="EE")

#convert the filtered OTU table from above to relative abundances and get the colnames back to match meta
rawabunds <- read_delim(here("Huus_Data","feature-table-for-R-2.txt"),"\t",comment = "# Constructed") %>% select(-c(1,247))
abunds <- relabund(data.frame(otu_table(Igf)))
colnames(abunds) <- colnames(rawabunds)
abunds <- abunds %>% select(which(colnames(abunds)%in%meta$samplename))

#subset meta to just those with sequencing and remove the EE condition to simplify comparisons
meta <- meta %>% slice(match(colnames(abunds),meta$samplename))

#seperate out the postive and negative and presort abundances
posabunds <- abunds %>% select(which(meta$Sort=="IgApos")) 
negabunds <- abunds %>% select(which(meta$Sort=="IgAneg"))
preabunds <- abunds %>% select(which(meta$Sort=="Pre"))

#convert the per-fraction ids to uniform sample names
sampname <- function(curnames){
  newnames <- gsub("-[A-z]*$","",curnames)
  return(newnames)
}
colnames(posabunds) <- sampname(colnames(posabunds))
colnames(negabunds) <- sampname(colnames(negabunds))
colnames(preabunds) <- sampname(colnames(preabunds))

#load in the pos and neg %s converting to sum to 1 and getting the sample name as above
fracsizes <- read_delim(here("Huus_Data","Table2-IgAposANDneg-cleaned.csv"),",") %>% rename("posper"=2,"negper"=3,"name"=4) %>% mutate(posper=posper/100,negper=negper/100) %>% add_column("sample"=sampname(.$name)) %>% filter(.$name%in%meta$samplename)

#get the list of samplenames with pos and neg abundances and fraction sizes
samps <- fracsizes$sample[which(fracsizes$sample%in%colnames(posabunds)&fracsizes$sample%in%colnames(negabunds))]

#subset all the tables to these samples and generate a meta table on the sample (not fraction) level
posabunds <- posabunds[,samps]
negabunds <- negabunds[,samps]
preabunds <- preabunds[,samps]

fracsizes <- fracsizes[match(samps,fracsizes$sample),]
possizes <- fracsizes$posper
names(possizes) <- fracsizes$sample
negsizes <- fracsizes$negper
names(negsizes) <- fracsizes$sample
sampmeta <- meta %>% add_column(sample=sampname(.$samplename)) %>% filter(Sort=="IgAneg") %>% slice(match(samps,.$sample)) %>% mutate(Age=as.numeric(gsub("Week","",.$Age)))

```

The dataset contained pre-sort, postive, and negative fraction sequencing for **`r {nrow(sampmeta)}`** samples. The experimental conditions are broken down as below, where week relates the age of the mouse in terms of the experimental time and diet the experimental condition (conventional or malnourished diet). All samples were from faecal pellets.

```{r tab, include=T}
kable(table(sampmeta$Age,sampmeta$Diet))

lowest <- min(min(posabunds[posabunds!=0]),min(negabunds[negabunds!=0]))
pseudo <- 5e-5
```

Filtering aiming to match that described in the paper led to retention of **`r nrow(posabunds)`** species features. The lowest observed abundance was **`r format(lowest, scientific=F)`** , so a pseudo count of **`r format(pseudo, scientific=F)`** will be used.

# Generate the Kau and Probability Ratio scores

```{r}
#generate a suitable pseudo counts
kau <- igascores(posabunds = posabunds, negabunds=negabunds, pseudo = pseudo, method="kau")
colnames(kau) <- colnames(posabunds)
probrat <- igascores(posabunds = posabunds, negabunds = negabunds, possizes = possizes, negsizes = negsizes, pseudo = pseudo, method="probratio")
colnames(probrat) <-  colnames(posabunds)
```

# Identifying taxa with differential IgA binding between conditions

As I am working with a limited subset of the previous data, I will focus on the intergroup comparisons and model age as continuous variable to increase power. A two-way ANOVA will be used modelling the effect of age, diet, and the interaction of age and diet on IgA binding of each taxon. This will be carried out using both the Kau index and the Probability Ratio. Significant cut-off is permissive (nominal p<0.05) given the small sample sizes (max 4 per group).

```{r}
#function to calculate smd with check for all zeros
ssmdfun <- function(v1,v2){
  if(length(v1)<2|length(v2)<2){
    return(NA)
  }
  comb <- c(v1,v2)
  if(length(which(comb!=0))>0){
    return(ssmd(v1,v2))
  }
  else{
    return(NA)
  }
}

#function to carry out anova and find any significance
igafun <- function(scores,met,thresh){
  scores <- as.numeric(scores)
  #carry out the staistical test if enough non na points (at least one sample in each diet and each age, min of 6 total samples)
  if(length(which(is.na(scores)))>0){
    met <- met[-which(is.na(scores)),]
    scores <- scores[-which(is.na(scores))]
  }
  res <- c(NA,NA,NA)
  names(res) <- c("Diet","Age:Diet","Any_Sig")
  w3 <- NA
  w5 <- NA
  w7 <- NA
  if(nrow(met)>6){
    if(all(dim(table(met$Age,met$Diet))==c(3,2))){
  an <- aov(scores~met$Age*met$Diet) %>% tidy()
  ps <- an$p.value[c(2,3)]
  res <- c(ps,ifelse(min(ps)<thresh,1,0))
  names(res) <- c("Diet","Age:Diet","Any_Sig")
  #generate normalised differences in MAL relative to CON at each time point
  w3 <- ssmdfun(scores[met$Age_Diet=="Week3MAL"],scores[met$Age_Diet=="Week3CON"])
  w5 <- ssmdfun(scores[met$Age_Diet=="Week5MAL"],scores[met$Age_Diet=="Week5CON"])
  w7 <- ssmdfun(scores[met$Age_Diet=="Week7MAL"],scores[met$Age_Diet=="Week7CON"])
  }
  }
  res <- c(res,"W3"=w3,"W5"=w5,"W7"=w7)
  return(res)
}

thresh=0.05

kresall <- unlist(apply(kau,1,igafun,met=sampmeta,thresh=thresh)) %>% t() %>% data.frame() %>% rownames_to_column("taxon") 
kres <- kresall %>% filter(Any_Sig==1)
prresall <- unlist(apply(probrat,1,igafun,met=sampmeta,thresh=thresh)) %>% t() %>% data.frame() %>% rownames_to_column("taxon")
prres <- prresall %>% filter(Any_Sig==1)

#function to plot significant hits
sigplot <- function(sigres,scores,met,name){
  #first plot of the standardised mean differences
  longdifs <- sigres %>% select("taxon","W3","W5","W7") %>% gather("Time","Score",-taxon) %>% mutate(taxon=gsub("^([A-z_]*;){3}","",taxon))
  meandifs <- longdifs %>% ggplot(aes(x=Time,y=Score,fill=taxon))+geom_col()+facet_wrap(.~taxon)+guides(fill=F)+ylab(paste0("Standardised Mean Difference ",name," MAL/CON"))+scale_fill_d3()
  #plot the actual scores
  scores <- scores[sigres$taxon,]
  longscores <- scores %>% rownames_to_column("taxon") %>% gather("Sample","Score",-taxon) %>%
    add_column(Time=met$Age[match(.$Sample,met$sample)],Diet=met$Diet[match(.$Sample,met$sample)]) %>%
    mutate(taxon=gsub("^([A-z_]*;){3}","",taxon))
  ldifs <- longscores %>% ggplot(aes(x=factor(Time),y=Score,fill=Diet))+geom_boxplot(outlier.shape = NA)+geom_point(aes(fill=Diet),position =position_dodge(width = 0.75),shape=21)+facet_wrap(.~taxon,scales="free")+ylab(name)+
    xlab("Age (Weeks)")+scale_fill_simpsons()
  return(list(mdifs=meandifs,ldifs=ldifs))
}

```

# Kau

```{r include=T, fig.width=10, fig.height=8}
kable(kres)

kps <- sigplot(kres,kau,sampmeta,"Kau")

kps$mdifs
```


```{r include=T, fig.width=10, fig.height=8}

kps$ldifs

```

# Probability-ratio

```{r include=T, fig.width=10, fig.height=8}
kable(prres)

prps <- sigplot(prres,probrat,sampmeta,"Probability Ratio")

prps$mdifs
```

```{r include=T, fig.width=10, fig.height=8}

prps$ldifs

```

# Comparison of effect sizes and p-values of significant associations between scoring methods

These plots compare the p-values for any comparison of diet or age*diet that is significant in either scoring method. Additionally, the difference in mean score between groups at all three time points is plotted for each of these comparisons.

```{r p_vs_eff, fig.width=10, fig.height=5, include=T}
#generate long format of all comparisons made between groups (age, diet, age*diet) and the effect sizes at each time point for each one
kreslong <- kresall %>% select(-Any_Sig) %>% gather("Comparison","P",-taxon,-W3,-W5,-W7) %>% gather("Week","EffectSize",-Comparison,-P,-taxon)
prlong <- prresall %>% select(-Any_Sig) %>% gather("Comparison","P",-taxon,-W3,-W5,-W7) %>% gather("Week","EffectSize",-Comparison,-P,-taxon)
colnames(kreslong) <- paste0("Kau_",colnames(kreslong))
colnames(prlong) <- paste0("PR_",colnames(prlong))
#plot
comb <- cbind(kreslong,prlong)
sigcomb <- comb %>% filter(PR_P<thresh|Kau_P<thresh) 
sigpplot <- sigcomb %>% ggplot(aes(y=PR_P,x=Kau_P))+geom_point()+geom_abline(slope=1,lty=2,col="red")+scale_y_continuous(limits=c(0,0.1))+scale_x_continuous(limits=c(0,0.1))+xlab("Kau Index p-value")+ylab("Probability Ratio p-value")
sigeffplot <- sigcomb %>% ggplot(aes(y=abs(PR_EffectSize),x=abs(Kau_EffectSize)))+geom_point()+geom_abline(slope=1,lty=2,col="red")+scale_y_continuous(limits=c(0,4))+scale_x_continuous(limits=c(0,4))+xlab("Absolute Kau Effect Size\n(SSMD MAL/CON)")+ylab("Absolute Probability Ratio\nEffect Size (SSMD MAL/CON)")

peff <- plot_grid(sigpplot,sigeffplot,nrow=1,align = "hv",rel_widths = c(0.9,1))

peff

```

# Comparing within group variation between the scores

This plot will look at the coefficient of variation within each taxa and each experimental group (age*diet) to see if there is reduced variance in score estimates in the Probability Ratio, as seen in the simulations.

```{r cov, fig.width=8, fig.height=8, include=T}
cv <- function(vals){
  return(sd(vals)/mean(vals))
}

kaucv <- kau %>% rownames_to_column("Taxon") %>% gather("Sample","Score",-Taxon) %>% add_column(Group=sampmeta$Age_Diet[match(.$Sample,sampmeta$sample)]) %>% group_by(Taxon,Group) %>% summarise(CV=cv(Score))
prcv <- probrat %>% rownames_to_column("Taxon") %>% gather("Sample","Score",-Taxon) %>% add_column(Group=sampmeta$Age_Diet[match(.$Sample,sampmeta$sample)]) %>% group_by(Taxon,Group) %>% summarise(CV=cv(Score))

cvs <- tibble(Group=kaucv$Group,Kau_CV=kaucv$CV,PR_CV=prcv$CV) %>% filter(!is.na(Kau_CV)&!is.na(PR_CV)) %>% mutate(Kau_CV=abs(Kau_CV),PR_CV=abs(PR_CV)) %>% gather("Method","CV",-Group) %>% mutate(Method=ifelse(Method=="Kau_CV","Kau Index","Probability Ratio"))
cvs %>% ggplot(aes(x=Method,y=CV))+geom_quasirandom(alpha=0.5,aes(col=Group))+geom_boxplot(outlier.shape=NA,fill=NA)+scale_y_log10()+stat_compare_means(method = "wilcox", comparisons = list(c(1,2)))+scale_color_d3()+
  theme(legend.position = "right",legend.direction = "vertical")
```





